\documentclass[12pt,letterpaper,titlepage,draft]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}

\renewcommand{\refname}{\section{References}}

\title{Thesis Proposal:\\
Selection and Mutation on Correlated\\
Neutral Fitness Landscapes}

\author{George Lesica\\
Department of Computer Science\\
University of Montana\\
Missoula, MT\\
\\
Advisor: Alden Wright}

\begin{document}
\maketitle

\section{Introduction}

\subsection{Biological Neutrality}

\paragraph{}
The theory of neutral evolution was first proposed by Kimura~\cite{Kimura1984}
to explain observed differences between evolution in populations of larger
organisms and evolution at the molecular level. His theory made two
predictions, as laid out by Hughes~\cite{Hughes2007}. First, ``that most
polymorphisms are selectively neutral and are maintained by genetic drift''.
Second, ``that most changes at the molecular level that are fixed over
evolutionary time are selectively neutral and are fixed by drift.''

\paragraph{}
Neutral evolution enables a population to ``explore'' genotype space via
neutral drift. Neutral mutation itself refers to mutations that have little or
no affect on the fitness of an organism, which is not to say that they have no
affect at all. In biology, this is the result of the many-to-one relationship
between genotypes and phenotypes~\cite{Newman1998}.

\paragraph{}
A population undergoing neutral evolution is likely to exhibit high genotypic
diversity. This presents opportunities for the population. Its members are able
to drift through genotypic space since selective pressure does not operate in
the context of neutral mutations. As a result, there is a greater likelihood
that some population member will ``discover'' a genotype that can easily mutate
into one that maps to a higher-fitness phenotype, elevating the entire
population when, and if, the mutation reaches fixation.

\subsection{Computational Modeling}

\paragraph{}
In computational modeling we can think of two genotypes that produce the same
phenotype (or which have equal fitness), each of which may be produced by
mutating the other in some small way, as two nodes in a neutral network. Many
such networks may exist for a particular genome, each corresponding to a
particular phenotype or fitness level. In some cases, each of these networks
will be small. In other cases, such networks may be expansive and may permeate
the genotype space. In the case where the neutral network for a given phenotype
(or fitness value) forms a single giant component (or comes close) it is said
to be a percolating neutral network.

\subsubsection{Fitness Landscapes}

\paragraph{}
A set of genotypes and their corresponding phenotypes or fitness levels make up
a fitness landscape, which is simply a useful visual metaphor. The intuition is
that higher ground within the landscape corresponds to higher fitness. Fitness
landscapes can be smooth or rugged. Smooth landscapes have lower peaks, in
other words there is less difference between fit and unfit genotypes.

\paragraph{}
Fitness landscapes can also be correlated or uncorrelated. A correlated
landscape exhibits auto-correlation. This is most naturally understood as the
expected correlation between successive fitness values during a random walk
across the landscape. Visually, a correlated landscape might look similar to
rolling hills, whereas an uncorrelated landscape would appear extremely jagged,
sometimes going from very low to very high in a short distance.

\subsubsection{Adaptive Walks}

\paragraph{}
An adaptive walk over a fitness landscape is the process of choosing a
genotype, then mutating it into a new, fitter genotype and repeating the
process. The fact that the new genotype is always fitter (or possibly at least
as fit) as the current genotype clearly differentiates an adaptive walk from a
random walk, which can also be useful to the study of fitness landscapes. It is
helpful to know that this process is often called ``hill climbing'', which
matches up with the landscape metaphor quite neatly.

\paragraph{}
There are several different methods for choosing the next genotype to occupy
during an adaptive walk. Some of the common ones are nicely summarized
in~\cite{Nowak2015}. These include fitness-dependent, where the choice is made
randomly, but fitter genotypes are given preference. Alternately, the choice
can be purely random. A greedy strategy chooses the fittest candidate, and
finally, a reluctant strategy chooses the least fit candidate that is fitter
than the current genotype.

\paragraph{}
Fitness landscapes have traditionally been seen as rugged, with many, more or
less independent, peaks and valleys pervading the space. Under these
conditions, it is possible, even likely, that a population undergoing an
adaptive walk will become ``trapped'' at a local optimum. Imagine attempting to
reach the summit of Mount Everest by starting where you are and continuously
walking toward a higher elevation. You will almost certainly end up at the top
of some hill or smaller mountain with nowhere to go.

\subsubsection{Neutral Networks}

\paragraph{}
Neutral landscapes, on the other hand, can be more accurately imagined as
mountain ranges, with ridges running along slopes joining together peaks,
saddles between the peaks themselves, and relatively flat valleys and meadows.
Adaptive walks on such landscapes are likely to result in higher fitness, as
demonstrated by Barnett~\cite{Barnett1998} or, put another way, populations are
less likely to become stuck at local optima~\cite{Newman1998}.

\paragraph{}
Another way to describe a neutral landscape is as a set of neutral basins, each
with some number of passageways that lead to other basins of higher or lower
fitness (a bit like a multi-level waterfall). A balance between selection and
deleterious mutation prevents the population from falling ``back'' into a lower
fitness basin~\cite{Crutchfield1999}.

\paragraph{}
We can think of exploring a fitness landscape by following neutral pathways
that may or may not form a percolating (or nearly so) network. Random mutation
drives this process. Eventually (or perhaps frequently), some individual will
mutate off of the neutral network and onto a neutral network with higher
fitness. If this mutation fixes in the population, then the population will
have adapted or evolved.

\section{NK Fitness Landscapes}

\paragraph{}
The NK fitness landscape, developed by Kauffman~\cite{Kauffman1993}, is a
fitness model that features tunable ruggedness. It derives its name from the
fact that an NK landscape is parameterized by two values: $N$, the number of
loci in each genotype, and $K$, the number of loci that are considered
epistatically linked to each locus.

\paragraph{}
The value of $K$ determines how rugged the landscape will be. Higher values
yield greater ruggedness. The special case where $K=0$ yields a landscape with
just a single peak.

\subsection{Construction and Evaluation}

\paragraph{}
An NK landscape is constructed by first choosing values for $N$ and $K$. Then,
for each locus in $N$, $K$ other loci are chosen at random. For each
combination of alleles for those $K+1$ loci (two alleles, 0 and 1, are
typically used for simplicity), a fitness value between zero and one is chosen.

\paragraph{}
Evaluation is done by computing the average fitness across the contributions
from each locus. The contribution for a single locus is computed using the
correct fitness based on the values of the $K+1$ relevant loci and the fitness
table computed when the landscape was created\footnote{In reality, because the
memory required to store such a table grows exponentially with $K$, fitness
values are more commonly generated on-demand. The result, however, is
unchanged.}.

\subsection{Variants}

\paragraph{}
One noteworthy, but perhaps not entirely obvious, implication of the NK design
is that the resultant landscape is almost certainly not neutral. The following
two variants adapt the general NK landscape to add neutrality.

\subsubsection{NKq Landscapes}

\subsubsection{NKp Landscapes}

\paragraph{}
The NKp landscape, due to Barnett~\cite{Barnett1998}, adds an additional
parameter $p$ which is assigned from the interval $\left[0,1\right]$. When
fitness values are assigned during creation of the landscape, fitness of
exactly 0 is chosen with probability $p$. This means that some loci will have
make no contribution to genotype fitness for certain values of the $K$ loci to
which they are linked.

\section{Hypotheses}

\subsection{Mutation-Selection Balance}

\paragraph{}
Gavrilets demonstrated~\cite{Gavrilets1997} that the percolation threshold for
an uncorrelated fitness landscape scales with the inverse of the number of
dimensions under consideration. The percolation threshold is the probability
that a given genotype is part of a particular neutral network (Gavrilets
limited fitness values to ``high'' and ``low'', so the neutral network in
question was the high fitness network).

\paragraph{}
In other words, as the number of dimensions goes up, it takes a smaller share
of the nodes to create a percolating neutral network. This makes sense because
each additional dimension increases the number of ``routes'' between two
points. For example, cross the Grand Canyon is quite an adventure in a car or
on foot, but if a helicopter is available, it becomes a fairly simple task.

\paragraph{}
We hypothesize that this holds for correlated landscapes as well, given an
appropriate balance between selection and mutation pressures.

\section{Methods}

\paragraph{}
To facilitate testing our hypotheses, we intend to create two pieces of
reusable software, both of which will be released as open source software.

\subsection{NK.jl Library}

\paragraph{}
We have opted to generalize our NK landscape
implementation\footnote{\url{https://github.com/glesica/NK.jl}} so that it may
be used by others in the future. We will provide it as a separate package under
a permissive license that will hopefully allow future work and improvement.  We
will use this library for our own experiments in addition to contributing it to
the community as open source software.

\paragraph{}
We intend to implement several NK variants including NKq and NKp neutral
landscapes. We will also provide implementations of several common techniques
such as adaptive walks and proportional selection from a population.

\subsection{CGP.jl Library}

\paragraph{}
Cartesian Genetic Programming (CGP) eschews the tree structures more commonly
associated with genetic programming in favor of a grid (hence ``Cartesian'') of
functions that can take as their inputs the outputs from any preceding
functions. Our CGP library\footnote{\url{https://github.com/glesica/CGP.jl}}
attempts to combine computation performance with ease-of-use.

\section{Experiments}

\subsection{Shifting Balance Experiment}

\paragraph{}
The first project will test the shifting balance theory proposed by
Wright~\cite{Wright1982}\cite{Wright1931}. We will enumerate the possible
genotypes for small NK and NKp landscapes, then compute the
prominence\footnote{Prominence is a concept from topography. The prominence of
a peak is defined to be the vertical distance between the peak and the
lowest contour line encircling it that has no higher summit.} and identify the
key col\footnote{The key col, also called the saddle point, of a peak is the
lowest point on the highest ridge between a peak and a higher peak that
defines its prominence.} for each fitness peak.

\paragraph{}
Ordinarily, a large population would be unlikely to move between fitness peaks,
even on a neutral landscape. Wright's theory attempted to resolve the apparent
conflict between theory and observation by suggesting that a subpopulation may
be able to sustain temporarily reduced fitness given a sufficiently high rate
of mutation and a sufficiently low rate of gene flow. Wright suggested that a
subpopulations might ``jump'' to a neighboring domain of attraction and climb
the corresponding peak. The beneficial mutation might then fix in the larger
population. We propose that this process is simpler and more likely to occur if
there exist neutral or nearly-neutral ridges between peaks.

\section{Analysis}

\bibliographystyle{acm}
\bibliography{fitness}{}
\end{document}
